Conservative Model Order Reduction for Fluid Flow

### 2018-10-05 backup.tex
updated version

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%% Combustor images %%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}
\centering
\begin{tikzpicture}[]
    \begin{axis}[width=12cm,
                 height=4cm,
                 ylabel = $r(m)$,
                 xlabel=$x(m)$,
                 yticklabel style={
                  /pgf/number format/fixed,
                  /pgf/number format/precision=5,
                  style = {font=\normalsize}
                  },
                 xticklabel style={
                  style = {font=\normalsize}
                  },
%                 label style={font=\large},
%                 legend style={font=\large},
%                 grid=both,
%                 ticks=both,
%                 width=0.7\linewidth, 
%                 height=0.5\linewidth,
                 minor x tick num=1,
                 minor y tick num=2,	
                 scaled y ticks = false,
%                 enlargelimits=false,
%                 samples = 100,
                 ymax = 0.04]
                 \addplot[color=black,style=solid,style=ultra thick]  table[x = x, y = radius] {./data/Combustor/radius_combustor.txt};
                \node[pin=100:{injector}] at (-0.05,0.01) {};
                \node[pin=93:{back-step}] at (0.001,0.015) {};
                \node[pin=90:{chamber}] at (0.2,0.023) {};
                \node[pin=80:{nozzle}] at (0.4,0.015) {};
    \end{axis}
\end{tikzpicture}
\caption{Geometry of quasi-1D CVRC model.}
\label{fig:radius}
\end{figure}

@@

\begin{table} [h]
	\centering
	\caption{CVRC operating conditions.}
	\centering
	\begin{tabular}{l l l}
		\toprule
		\centering
		Parameter & Unit & Value \\
		\midrule
		Fuel mass flow rate, $\dot{m}_{f}$ & kg/s & 0.027   \\
		Fuel temperature, $T_{f}$ & K & 300   \\
		Oxidizer mass flow rate, $\dot{m}_{ox}$ & kg/s & 0.32   \\
		Oxidizer temperature, $T_{ox}$ & K & 1030   \\
		$O_2$ mass fraction in oxidizer, $Y_{O_2}$ & -- & 42.4\%   \\
		$H_2O$ mass fraction in oxidizer, $Y_{H_2O}$ & -- & 57.6\%   \\
		Mean chamber pressure & MPa & 1.34 \\
		Equivalence ratio, $E_r$ & -- & 0.8 \\
		\bottomrule
		\label{tab:operating-conditions}
	\end{tabular} 
\end{table}
For the combustion, we consider the one-step reaction model
\begin{equation*}\label{eq:combustion}
CH_4 + 2O_2 \rightarrow CO_2 + 2H_2O
\end{equation*}
We assume that the fuel reacts instantaneously to form products, allowing us to neglect intermediate species and finite reaction rates. As the equivalence ratio is less than one, there is oxidizer left after the combustion. Therefore, only two species need to be considered: oxidizer and combustion products.

@@

%\begin{figure}[t]
%\centering
%\begin{subfigure}[]{0.47\linewidth}
%	\includegraphics[width=\textwidth]{./pic/init_pressure}
%	\caption{} \label{fig:5p2.2a}
%\end{subfigure}
%\begin{subfigure}[]{0.47\linewidth}
%	\includegraphics[width=\textwidth]{./pic/init_osc}
%	\caption{} \label{fig:5p2.2b}
%\end{subfigure} \\
%\begin{subfigure}[]{0.47\linewidth}
%	\includegraphics[width=\textwidth]{./pic/p_error}
%	\caption{} \label{fig:5p2.2c}
%\end{subfigure}
%\begin{subfigure}[]{0.47\linewidth}
%	\includegraphics[width=\textwidth]{./pic/rom_freq}
%	\caption{} \label{fig:5p2.2d}
%\end{subfigure}
%\caption{(a) Pressure profile of the steady state. (b) Oscillatory mode of pressure located at $x=0.36$ for the unsteady flow. (c) Relative error between the high-fidelity and approximated pressure. (d) Approximation of the oscillations. }
%\label{fig:5p2.2}
%\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%% Combustor %%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}
  \centering
  % Pressure steady
  \begin{subfigure}[]{0.48\linewidth}
  \begin{tikzpicture}[scale=0.55]
    \begin{axis}[ylabel = pressure $p$,
                 xlabel=$x$,
                 label style={font=\Large},
                 legend pos=north east,
                 %legend entries={$k=102$ Adv,$k=201$ Adv,$k=102$ Div,$k=201$ Div,$k=102$ Skew,$k=201$ Skew},
                 legend style={font=\large},
                 grid=both,
                 ticks=both,
                 width=1.4\linewidth, 
                 height=1.0\linewidth,
                 minor x tick num=1,
                 minor y tick num=2,	
                 %yticklabel style={/pgf/number format/.cd,fixed,precision=9},
                 scaled x ticks = true,
                 enlargelimits=false,
                 scale only axis,
%                 ymin=0,
                 ymax = 1500000,
                 samples = 100]
                 \addplot[color=blue,style=solid,style=ultra thick]  table[x = x, y = pressure] {./data/Combustor/pressure_steady_state.txt};
    \end{axis}%  
  \end{tikzpicture}
  \caption{} \label{fig:5p2.2a}
  \label{steady_state}
  \end{subfigure}\hfill% 
  % Pressure oscilaltions
  \begin{subfigure}[]{0.48\linewidth}
  \begin{tikzpicture}[scale=0.55]
    \begin{axis}[ylabel = pressure $p$,
                 xlabel=$t$,
                 label style={font=\Large},
                 legend pos=north east,
                 %legend entries={$k=102$ Adv,$k=201$ Adv,$k=102$ Div,$k=201$ Div,$k=102$ Skew,$k=201$ Skew},
                 legend style={font=\large},
                 grid=both,
                 ticks=both,
                 width=1.4\linewidth, 
                 height=1.0\linewidth,
                 minor x tick num=1,
                 minor y tick num=2,	
                 %yticklabel style={/pgf/number format/.cd,fixed,precision=9},
                 scaled x ticks = true,
                 scaled y ticks = true,
                 enlargelimits=false,
                 scale only axis,
                 ymin=0,
%                 samples = 100,
                 xmax = 0.06,
                 ymin = 1360000,
                 ymax = 1460000,
                 tick label style={/pgf/number format/fixed} ]
                 \addplot[color=blue,style=solid,style=thick]  table[x = t, y = pressure] {./data/Combustor/ciao.txt};
    \end{axis}%  
  \end{tikzpicture}
  \caption{} \label{fig:5p2.2b}
  \label{oscill_pressure}
  \end{subfigure}

% Total momentum
  \begin{subfigure}[]{0.48\linewidth}
  \begin{tikzpicture}[scale=0.55]
    \begin{semilogyaxis}[ylabel = $\| p-\tilde{p}\|_2$,
                 xlabel=$t$,
                 label style={font=\Large},
                 legend pos=south east,
                 legend entries={$k=200$,$k=300$,$k=300$},
                 legend style={font=\large},
                 grid=both,
                 ticks=both,
                 width=1.4\linewidth, 
                 height=1.0\linewidth,
                 minor x tick num=1,
                 minor y tick num=2,	
                 %yticklabel style={/pgf/number format/.cd,fixed,precision=9},
                 scaled x ticks = true,
                 scaled y ticks = true,
                 enlargelimits=false,
                 scale only axis,
                 samples = 300,
                 ymin = 0.000001,
                 ymax = 0.0001]
                 \addplot[color=red,style=solid,style=thick]  table[x = t, y = error] {./data/Combustor/error_200.txt};
                 \addplot[color=blue,style=solid,style=thick]  table[x = t, y = error] {./data/Combustor/error_300.txt};
                 \addplot[color=black!50!green,style=solid,style=thick]  table[x = t, y = error] {./data/Combustor/error_400.txt};
    \end{semilogyaxis}%  
  \end{tikzpicture}
  \caption{} \label{fig:5p2.2c}
  \label{error_growth_comb}
  \end{subfigure}\hfill% 
  % Pressure reduction  
  \begin{subfigure}[]{0.48\linewidth}
  \begin{tikzpicture}[scale=0.55]
    \begin{axis}[ylabel = pressure $p$,
                 xlabel=$t$,
                 label style={font=\Large},
                 legend pos=north east,
                 legend entries={High fidelity,  POD with $k=200$},
                 legend style={font=\large},
                 grid=both,
                 ticks=both,
                 width=1.4\linewidth, 
                 height=1.0\linewidth,
                 minor x tick num=1,
                 minor y tick num=2,	
                 %yticklabel style={/pgf/number format/.cd,fixed,precision=9},
                 scaled x ticks = true,
                 enlargelimits=false,
                 scale only axis,
                 xmin = 0.9,
                 xmax = 0.98,
                 ymin = 1390000,
                 ymax = 1450000,
                 tick label style={/pgf/number format/fixed} ]
                 \addplot[color=green,style=solid,style=ultra thick]  table[x = t, y = pressure] {./data/Combustor/full.txt};
                 \addplot[color=black,style=dashed,style=thick]  table[x = t, y = pressure] {./data/Combustor/red.txt};
    \end{axis}%  
  \end{tikzpicture}
  \caption{} \label{fig:5p2.2d}
  \label{oscill_focus}
  \end{subfigure}
  \caption{(\protect\subref{steady_state}) Pressure profile of the steady state.  (\protect\subref{oscill_pressure}) Oscillatory mode of pressure located at $x=0.36$ for the unsteady flow. (\protect\subref{error_growth_comb}) Relative error between the high-fidelity and approximated pressure. (\protect\subref{oscill_focus}) Approximation of the oscillations.} 
  \label{fig:5p2.2}
\end{figure}

>>>>

\begin{figure}
	\centering
	\includegraphics[width=0.8\linewidth]{pic/radius}
	\caption{Geometry of quasi-1D CVRC model.}
	\label{fig:radius}
\end{figure}

@@

\begin{table} [h]
	\centering
	\caption{CVRC operating conditions.}
	\centering
	\begin{tabular}{l l l}
		\toprule
		\centering
		Parameter & Unit & Value \\
		\midrule
		Fuel mass flow rate, $\dot{m}_{f}$ & kg/s & 0.027   \\
		Fuel temperature, $T_{f}$ & K & 300   \\
		Oxidizer mass flow rate, $\dot{m}_{ox}$ & kg/s & 0.32   \\
		Oxidizer temperature, $T_{ox}$ & K & 1030   \\
		$O_2$ mass fraction in oxidizer, $Y_{O_2}$ & -- & 42.4\%   \\
		$H_2O$ mass fraction in oxidizer, $Y_{H_2O}$ & -- & 57.6\%   \\
		Mean chamber pressure & MPa & 1.34 \\
		Equivalence ratio, $E_r$ & -- & 0.8 \\
		\bottomrule
		\label{tab:operating-conditions}
	\end{tabular} 
\end{table}

For the combustion, we consider the one-step reaction model
\begin{equation*}\label{eq:combustion}
CH_4 + 2O_2 \rightarrow CO_2 + 2H_2O
\end{equation*}
We assume that the fuel reacts instantaneously to form products, allowing us to neglect intermediate species and finite reaction rates. As the equivalence ratio is less than one, there is oxidizer left after the combustion. Therefore, only two species need to be considered: oxidizer and combustion products.

@@

\begin{figure}[t]
\centering
\begin{subfigure}[]{0.47\linewidth}
	\includegraphics[width=\textwidth]{./pic/init_pressure}
	\caption{} \label{fig:5p2.2a}
\end{subfigure}
\begin{subfigure}[]{0.47\linewidth}
	\includegraphics[width=\textwidth]{./pic/init_osc}
	\caption{} \label{fig:5p2.2b}
\end{subfigure} \\
\begin{subfigure}[]{0.47\linewidth}
	\includegraphics[width=\textwidth]{./pic/p_error}
	\caption{} \label{fig:5p2.2c}
\end{subfigure}
\begin{subfigure}[]{0.47\linewidth}
	\includegraphics[width=\textwidth]{./pic/rom_freq}
	\caption{} \label{fig:5p2.2d}
\end{subfigure}
\caption{(a) Pressure profile of the steady state. (b) Oscillatory mode of pressure located at $x=0.36$ for the unsteady flow. (c) Relative error between the high-fidelity and approximated pressure. (d) Approximation of the oscillations. }
\label{fig:5p2.2}
\end{figure}

### 2018-10-05 paper.tex
Update remaining figures Nicolo
updated version

@@

\title*{Conservative Model Order Reduction for Fluid Flow}
% Use \titlerunning{Short Title} for an abbreviated version of
% your contribution title if the original one is too long
\author{Babak Maboudi Afkham, Nicol\`o Ripamonti, Qian Wang, Jan S. Hesthaven and Karen E. Willcox}
% Use \authorrunning{Short Title} for an abbreviated version of
% your contribution title if the original one is too long
\institute{Name of First Author \at Name, Address of Institute, \email{name@email.address}
\and Name of Second Author \at Name, Address of Institute \email{name@email.address}}
%
% Use the package "url.sty" to avoid
% problems with special characters
% used in your e-mail or web address
%
\maketitle

>>>>

\pgfplotsset{every tick label/.append style={font=\Large}}

@@

\title*{Conservative Model Order Reduction for Fluid Flow}
% Use \titlerunning{Short Title} for an abbreviated version of
% your contribution title if the original one is too long
\author{Babak Maboudi Afkham, Nicol\`o Ripamonti, Qian Wang, and Jan S. Hesthaven}
% Use \authorrunning{Short Title} for an abbreviated version of
% your contribution title if the original one is too long
\institute{B. Maboudi Afkham, N. Ripamonti, Q. Wang, J.S. Hesthaven \at EPFL, Lausanne, Switzerland \email{babak.maboudi@epfl.ch}}
%
% Use the package "url.sty" to avoid
% problems with special characters
% used in your e-mail or web address
%
\maketitle

### 2018-10-05 sections/1.intro.tex
updated version

Skew-symmetric formulation of fluid flows construct a skew-symmetric differential operator, acting on the momentum vector field, that ensures conservation of quadratic invariants, such as energy. Combined wirh centered time and space discretization schemes, typically a finite differences discretization method, they recover time-symmetries of a fluid at the discrete level. Such discretization schemes are studies comprehensively over the past few decades and can be found in the works of \cite{morinishi2010skew,morinishi1998fully,desjardins2008high,tadmor1984skew,reiss2014conservative,reiss2014conservative} and the references therein.

>>>>

Skew-symmetric formulation of fluid flows constructs a skew-symmetric differential operator, acting on the momentum vector field, that ensures conservation of quadratic invariants, such as energy. Combined wirh centered time and space discretization schemes, typically a finite differences discretization method, they recover time-symmetries of a fluid at the discrete level. Such discretization schemes are studies comprehensively over the past few decades and can be found in the works of \cite{morinishi2010skew,morinishi1998fully,desjardins2008high,tadmor1984skew,reiss2014conservative} and the references therein.

### 2018-10-05 sections/4.mor_skew.tex
updated version

	&\frac{1}{\gamma -1} \frac{d}{dt} \mathbf p^r + \frac{\gamma}{\gamma -1} V_{\mathbf p}^T D U V_{\mathbf p} \mathbf p^r - V_{\mathbf p}^T UD V_{\mathbf p} \mathbf p^r = - V_{\mathbf p}^T UDT + V_{\mathbf p}^T D(UT - \mathbf \phi).

@@

Note that in \eqref{eq:4.4}, dependency of $T$ on $V_{\mathbf r \mathbf u , \mathbf u}$ is not shown for abbreviation. In \eqref{eq:4.3} and \eqref{eq:4.4}, $D_i$ is always multiplied from the left with a basis matrix or a diagonal matrix. Therefore, the telescoping sum, discussed in Condition 2 in \Cref{sec:skew.1}, cannot be used to show conservation of mass and momentum. However, POD preserves linear properties of snapshots. To demonstrate this, let the overscript ``\textasciitilde'' denote the representation of a reduced variable in the high fidelity space. A reduced variable, e.g. density, can be represented as a linear combination of some snapshots as $\mathbf r \approx \tilde{\mathbf r} = \sum_{i=1}^k c_i \mathbf r_i$, for some snapshots $\mathbf r_i$ and some coefficients $c_i \in \mathbb R$, for $i=1,\dots,k$. Conservation of mass, evaluated by $\tilde{\mathbf r}$, reads

@@

In the missing steps in the last line, skew-symmetry of $S_{\mathbf r \mathbf u}^r$ is used. Note, that only the reduced pressure and the viscous term contribute to the evolution of kinetic energy. Furthermore, the quantity $ \frac 1 2 {\mathbf u^r}^T \{ R \mathbf u\}^r$ is the kinetic energy associated with the reduced system \eqref{eq:4.4}, approximating the kinetic energy of the high-fidelity system \eqref{eq:3.21} and is a quadratic form with respect to the reduced variables. Conservation of kinetic energy for \eqref{eq:4.3} follows similarly. It is  straight-forward to check that

>>>>

	&\frac{1}{\gamma -1} \frac{d}{dt} \mathbf p^r + \frac{\gamma}{\gamma -1} V_{\mathbf p}^T D U V_{\mathbf p} \mathbf p^r - V_{\mathbf p}^T UD V_{\mathbf p} \mathbf p^r = - V_{\mathbf p}^T UDT + V_{\mathbf p}^T D(UT - \mathbf \phi),

@@

respectively. Note that in \eqref{eq:4.4}, dependency of $T$ on $V_{\mathbf r \mathbf u , \mathbf u}$ is not shown for abbreviation. In \eqref{eq:4.3} and \eqref{eq:4.4}, $D_i$ is always multiplied from the left with a basis matrix or a diagonal matrix. Therefore, the telescoping sum, discussed in Condition 2 in \Cref{sec:skew.1}, cannot be used to show conservation of mass and momentum. However, POD preserves linear properties of snapshots. To demonstrate this, let the overscript ``\textasciitilde'' denote the representation of a reduced variable in the high-fidelity space. An approximated variable, e.g. density, can be represented as a linear combination of some snapshots as $\mathbf r \approx \tilde{\mathbf r} = \sum_{i=1}^k c_i \mathbf r_i$, for some snapshots $\mathbf r_i$ and some coefficients $c_i \in \mathbb R$, for $i=1,\dots,k$. Conservation of mass, evaluated by $\tilde{\mathbf r}$, reads

@@

In the missing steps in the last line, skew-symmetry of $S_{\mathbf r \mathbf u}^r$ is used. Note, that only the reduced pressure and the viscous term contribute to the evolution of kinetic energy. Furthermore, the quantity $ \frac 1 2 {\mathbf u^r}^T \{ R \mathbf u\}^r$ is the kinetic energy associated with the reduced system \eqref{eq:4.4}, approximating the kinetic energy of the high-fidelity system \eqref{eq:3.21}, and is a quadratic form with respect to the reduced variables. Conservation of kinetic energy for \eqref{eq:4.3} follows similarly. It is  straight-forward to check that

### 2018-10-05 sections/5.res.tex
updated version

\subsection{1D Shock problem}
In this section we study the 1-dimensional compressible Euler problem, \eqref{eq:3.18} without viscous terms, with a steady state discontinuous solution. This is in preparation for \Cref{sec:5.res.4}, where development and propagation of shock waves is discussed. Here we asses how the skew-symmetric form of \eqref{eq:3.18} can recover moving discontinuities at the level of the reduced system. Consider a periodic boundary conditions on $\Omega = [0,1]$ with the initial condition
\begin{equation*}
\begin{cases}
& \mathbf{r} = 0.5+0.2 \cos(2\pi x),\\
& \mathbf{u} = 1.5,\\
& \mathbf{p} = 0.5+0.2 \sin(2\pi x).\\
\end{cases}
\end{equation*}
The domain is discretized into $N=2000$ nodes and a centered finite differences scheme are used to assemble the discrete Euler equation in skew-symmetric form, as discussed in \Cref{sec:skew.3}.

@@

		\Cref{fig:energy_pres_1D} shows the evolution of conserved quantities for the high-fidelity and reduced system. Here, the high-fidelity model is also considered in the divergence and advective form in addition to the skew-symmetric form. It is clear than when the reduced systems is not in skew-symmetric form, it violates conservation of mass, momentum, and energy. Even while the high-fidelity systems in divergence and advective forms are stable, the constructed reduced system is unstable, independent on the number of basis vectors. On the other hand, the skew-symmetric form yields a stable and conservative reduce system. Note that the loss in the energy associated with the skew-symmetric form, illustrated in \Cref{conservation_1D_focus}, is due to the application of an artificial viscosity.

@@

As discussed in \Cref{sec:mor_skew}, the DEIM approximation needed for efficient evaluation of the nonlinear components of \eqref{eq:3.18}, can affect the conservation properties of the skew-symmetric form. \Cref{fig:singular_values_decay_1D} shows the decay of the singular values of the nonlinear snapshots. The decay of these snapshots is significantly slower than the temporal snapshots of \eqref{eq:3.18}. This indicates that to maintain the accuracy of the reduced system, the DEIM basis should be chosen richer than the POD basis. \Cref{error_DEIM} and \Cref{energy_DEIM} present the error and the conservation of total energy when the DEIM is used to approximate the nonlinear term. The conservation of energy is recovered once DEIM approximates the nonlinear terms with enough accuracy. While straightforward POD reduction does not offer a significative speed up if compared to the solution of the original high-dimensional nonlinear system of equations, with the DEIM treatment of nonlinearity solvers of the reduced models are of the order of tens faster than high fidelity solvers.

>>>>

\subsection{1D Shock problem}
In this section we study the 1-dimensional compressible Euler problem, \eqref{eq:3.18} without viscous terms, with a steady state discontinuous solution. This is in preparation for \Cref{sec:5.res.4}, where development and propagation of shock waves is discussed. Here we asses how the skew-symmetric form of \eqref{eq:3.18} can recover moving discontinuities at the level of the reduced system. Consider a periodic boundary conditions on $\Omega = [0,1]$ with the initial condition
\begin{equation*}
\begin{cases}
& \mathbf{r} = 0.5+0.2 \cos(2\pi x),\\
& \mathbf{u} = 1.5,\\
& \mathbf{p} = 0.5+0.2 \sin(2\pi x).\\
\end{cases}
\end{equation*}
The domain is discretized into $N=2000$ nodes and a centered finite differences scheme is used to assemble the discrete Euler equation in skew-symmetric form, as discussed in \Cref{sec:skew.3}.

@@

		\Cref{fig:energy_pres_1D} shows the evolution of conserved quantities for the high-fidelity and reduced system. Here, the high-fidelity model is also considered in the divergence and advective form in addition to the skew-symmetric form. It is clear that when the reduced systems is not in skew-symmetric form, it violates conservation of mass, momentum, and energy. Even while the high-fidelity systems in divergence and advective forms are stable, the constructed reduced system is unstable, independent on the number of basis vectors. On the other hand, the skew-symmetric form yields a stable and conservative reduce system. Note that the loss in the energy associated with the skew-symmetric form, illustrated in \Cref{conservation_1D_focus}, is due to the application of an artificial viscosity.

@@

As discussed in \Cref{sec:mor_skew}, the DEIM approximation needed for an efficient evaluation of the nonlinear components of \eqref{eq:3.18}, can affect the conservation properties of the skew-symmetric form. \Cref{fig:singular_values_decay_1D} shows the decay of the singular values of the nonlinear snapshots. The decay of these snapshots is significantly slower than the temporal snapshots of \eqref{eq:3.18}. This indicates that to maintain the accuracy of the reduced system, the DEIM basis should be chosen richer than the POD basis. \Cref{error_DEIM} and \Cref{energy_DEIM} present the error and the conservation of total energy when the DEIM is used to approximate the nonlinear term. The conservation of energy is recovered once DEIM approximates the nonlinear terms with enough accuracy. 
In this numerical experiment, evaluation of the nonlinear terms in \eqref{eq:3.18} using the DEIM is ten times faster than the high-fidelity evaluation.
